Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-29 Manuscript/Figures/figure_legends.Rmd
editted manuscript and legends

**Figure 6. Comparing annotation-specific random forest models implemented on SMOTE re-sampled data with distance-type predictors for GM12878.** MCC performances were compared for models using specific groups of annotations as the feature space including: histone modifications (HM; red), chromatin state segmentations (CSS; green), transcription factor binding sites (TFBS; blue), and a feature space containing all 3 types combined (combined; purple). Each bar represents a model ran through the ensemble framework, with re-sampling technique and predictor-type characteristics fixed to SMOTE and distance-types. The error bars represent $\pm$ 1 standard deviation from the mean MCC across each chromosome.

**Figure 7. Predictive importance.** A heatmap illustrating the importance of the top 20 most predictive transcription factors for random forest models implemented on SMOTE resampled data with distance-type predictors for GM12878. Importances were sorted according to the 10 kb resolution and rescaled to values between 0 and 100. Transcription factors with higher values of predictive importance are denoted in dark red.

**Figure 8. Comparing $log_{2}$ distance profiles.** Boxplots illustrating the distribution of distances around 10 kb regions on the GM12878 genome that either overlapped with a flanked TAD boundary (Yes) or did not (No), for the most influential TFBS including SMC3, CTCF, RAD21, ZNF143, and YY1sc281. P-values obtained from Wilcoxon Rank Sum tests are provided assessing the significance [**** = p < 0.00001].

<!--
**Figure 7. Comparing $log_{2}$ distance profiles of top predictive transcription factor binding sites by bin class for 10 kb resolution, on GM12878.** Violin plots (top) and plots of differences in quantiles (bottom) between genomic bin class for 4 of the top predictive TFBS determined by the best performing model, including: SMC3 (A), ZNF143 (B), CTCF (C), and RAD21 (D). For the plots of differences in quantiles, the reference group are the genomic bins that did not overlap with a flanked TAD boundary. That is, quantile differences above 0 indicate that bins that overlapped with flanked TAD boundaries exhibited smaller quantiles.
-->

>>>>

**Figure 6. Optimally built TFBS-specific models outperformed other annotation-specific models across resolution for GM12878** Performances were compared for random forest models built using specific groups of annotations as the feature space including: histone modifications (HM; red), chromatin state segmentations (CSS; green), transcription factor binding sites (TFBS; blue), and a feature space containing all 3 types combined (combined; purple). Each bar represents a model ran through the ensemble framework, with re-sampling technique and predictor-type characteristics fixed to SMOTE and distance-types. The error bars represent $\pm$ 1 standard deviation from the mean MCC across each chromosome.

**Figure 7. Top 20 most predictive transcription factor binding sites sorted by 10 kb for GM12878.** A heatmap illustrating the importance of the key cell line specific transcription factors for optimally built random forest models across resolution. Importances were sorted according to the 10 kb resolution and rescaled to values between 0 and 100. Transcription factors with higher values of predictive importance are denoted in dark red.

**Figure 8. Top predictive transcription factor binding sites were significantly closer to bins that overlapped with 10 kb flanked TAD boundaries on GM12878** Boxplots illustrating the distribution of distances (on $log_{2}$ scale) around 10 kb regions on the linear genome that either overlapped with a flanked TAD boundary (Yes) or did not (No), for the most influential cell line specific TFBS including SMC3, CTCF, RAD21, ZNF143, and YY1sc281. P-values obtained from Wilcoxon Rank Sum tests are provided assessing the significance [$NS$ = p > 0.05; $*$ = 0.01 < p $\le$ 0.05; $**$ = 0.001 < p $\le$ 0.01; $***$ = 0.0001 < p $\le$ 0.001; $****$ = p $\le$ 0.00001].

### 2019-05-29 Manuscript/Supplementary/supplementary_legends.Rmd
editted manuscript and legends

**Supplementary Figure 5. Comparing annotation-specific random forest models implemented on SMOTE re-sampled data with distance-type predictors for K562.** MCC performances were compared for models using specific groups of annotations as the feature space including: histone modifications (HM; red), chromatin state segmentations (CSS; green), transcription factor binding sites (TFBS; blue), and a feature space containing all 3 types combined (combined; purple). Each bar represents a model ran through the ensemble framework, with re-sampling technique and predictor-type characteristics fixed to SMOTE and distance-types. The error bars represent $\pm$ 1 standard deviation from the mean MCC across each chromosome.

**Supplementary Figure 6. Predictive importance** A heatmap illustrating the importance of the top 20 most predictive transcription factors for random forest models implemented on SMOTE resampled data with distance-type predictors for K562. Importances were sorted according to the 10 kb resolution and rescaled to values between 0 and 100. Transcription factors with higher values of predictive importance are denoted in dark red.

**Figure 9. Comparing $log_{2}$ distance profiles.** Boxplots illustrating the distribution of distances around (A) 25 kb, (B) 50 kb, and (C) 100 kb regions on the GM12878 genome that either overlapped with a flanked TAD boundary (Yes) or did not (No), for the most influential TFBS including SMC3, CTCF, RAD21, ZNF143, and YY1sc281. P-values obtained from Wilcoxon Rank Sum tests are provided assessing the significance [**** = p < 0.00001].

**Figure 10. Comparing $log_{2}$ distance profiles.** Boxplots illustrating the distribution of distances around (A) 10 kb, (B) 25 kb, (C) 50 kb, (D) 100 kb regions on the GM12878 genome that either overlapped with a flanked TAD boundary (Yes) or did not (No), for the most influential TFBS including SMC3, CTCF, RAD21, and ZNF143. P-values obtained from Wilcoxon Rank Sum tests are provided assessing the significance [**** = p < 0.00001].

>>>>

**Supplementary Figure 5. Optimally built TFBS-specific models outperformed other annotation-specific models across resolution for K562** Performances were compared for random forest models built using specific groups of annotations as the feature space including: histone modifications (HM; red), chromatin state segmentations (CSS; green), transcription factor binding sites (TFBS; blue), and a feature space containing all 3 types combined (combined; purple). Each bar represents a model ran through the ensemble framework, with re-sampling technique and predictor-type characteristics fixed to SMOTE and distance-types. The error bars represent $\pm$ 1 standard deviation from the mean MCC across each chromosome.

**Supplementary Figure 6. Top 20 most predictive transcription factor binding sites sorted by 10 kb for K562.** A heatmap illustrating the importance of the key cell line specific transcription factors for optimally built random forest models across resolution. Importances were sorted according to the 10 kb resolution and rescaled to values between 0 and 100. Transcription factors with higher values of predictive importance are denoted in dark red.

**Supplementary Figure 9. Top predictive transcription factor binding sites were significantly closer to bins that overlapped with varying units of flanked TAD boundaries on GM12878** Boxplots illustrating the distribution of distances (on $log_{2}$ scale) around (A) 25 kb, (B) 50 kb, and (C) 100 kb regions on the linear genome that either overlapped with a flanked TAD boundary (Yes) or did not (No), for the most influential cell line specific TFBS including SMC3, CTCF, RAD21, ZNF143, and YY1sc281. P-values obtained from Wilcoxon Rank Sum tests are provided assessing the significance [$NS$ = p > 0.05; $*$ = 0.01 < p $\le$ 0.05; $**$ = 0.001 < p $\le$ 0.01; $***$ = 0.0001 < p $\le$ 0.001; $****$ = p $\le$ 0.00001].

**Supplementary Figure 10. Top predictive transcription factor binding sites were significantly closer to bins that overlapped with varying units of flanked TAD boundaries on K562** Boxplots illustrating the distribution of distances (on $log_{2}$ scale) around (A) 10 kb, (B) 25 kb, (C) 50 kb, and (D) 100 kb regions on the linear genome that either overlapped with a flanked TAD boundary (Yes) or did not (No), for the most influential cell line specific TFBS including SMC3, CTCF, RAD21, and ZNF143. P-values obtained from Wilcoxon Rank Sum tests are provided assessing the significance [$NS$ = p > 0.05; $*$ = 0.01 < p $\le$ 0.05; $**$ = 0.001 < p $\le$ 0.01; $***$ = 0.0001 < p $\le$ 0.001; $****$ = p $\le$ 0.00001].

### 2019-05-29 Manuscript/manuscript.Rmd
editted manuscript and legends

We then investigated the spatial patterns of the most predictive TFBS around TAD boudaries by characterizing the distance profiles between the classes of genomic bins. For GM12878, at 10 kb resolution, we found that SMC3, CTCF, RAD21, ZNF143, and YY1sc281 were all statistically significantly closer to bins that overlapped with flanked TAD boudaries than those that did not (Figure 8). Although still statistically significantly different, the distance profiles between genomic class were found to be much more similar at lower resolutions (Supplementary 9). For the K562 cell line, we reported the distance profiles of the same transcription factor binding sites, excluding the YY1sc281, due to its unavailability in the UCSC Genome Browser. These cell line specific distance profiles were again found to be significantly different between genomic class, at all resolutions (Supplementary 10).

>>>>

We then investigated the spatial patterns of the most predictive TFBS around TAD boudaries by characterizing the distance profiles between the classes of genomic bins. For GM12878, at 10 kb resolution, we found that SMC3, CTCF, RAD21, ZNF143, and YY1sc281 were all statistically significantly closer to bins that overlapped with flanked TAD boudaries than those that did not (Figure 8). Although still statistically significantly different, the distance profiles between genomic class were found to be much more similar at lower resolutions (Supplementary Figure 9). For the K562 cell line, we reported the distance profiles of the same transcription factor binding sites, excluding the YY1sc281, due to its unavailability in the UCSC Genome Browser. These cell line specific distance profiles were again found to be significantly different between genomic class, at all resolutions (Supplementary Figure 10).
